# Homework Example

### Dan Birman
### 12/23/14 v2

## Function Load

```{r}
library(ggplot2)
library(lme4)
```

## Background

```{r}
f_data = read.csv("http://web.stanford.edu/class/psych252/_downloads/families.csv")
str(f_data)
```

Using `f_data` from:
Data from a study of 68 companies, examining relationships between the quality of family-friendly programs at each company, the percentage of employees with families who use these programs, and employee satisfaction (all continuous variables).

Fields:

famprog: the amount of family-friendly programs from (1 = Nothing at all to 9 = Amazing family-friendliness)
perfam: the percentage of employees with families in the organization (from 0% to 100%)
empastis: the average rating of employee satisfaction (1 = Extremely unsatisfied to 7 = Extremely satisfied)

*lots of additional background, motivations, conceptual details, etc, could go here*

## Questions

Develop the solutions to the following questions:

  **(1)** Describe the effects of famprog and perfam on empsatis: Do family-friendly programs improve employee satisfaction overall? Does the percentage of employees with families impact employee satisfaction, and how might these be related?\
  **(2)** Interpret your answer to Question (1) by testing the effect of family programs on employee satisfaction for companies who have the average number of employees with families, for companies who have relatively few employees with families (-1SD), and for companies who have relatively a lot of employees with families (+1SD).\
  **(3)** What do you conclude?
  
An excellent solution might include data analysis, graphical visualizations, and speculation about the underlying processes. Remember that your results should explain in detail your thought process and your output: why did you choose the analysis you chose and what do the results indicate? APA format should be used for all of your statistical results. In addition, keep in mind that it is difficult to understand what statistics reflect (i.e. +1 SD) without context, give an indication of actual values in your results.

## Example Code

You *may* need to use and understand the output from the following functions. (Note: not the right example code for this particular question, but you get the idea)

```{r}
dataset = data.frame(dependentVar=rnorm(100,0,25),independentVar=rnorm(100,5,20),subjectVar=c(1:50,1:50))

# Use code similar to this to plot the data and get a visual of what is happening
ggplot(data=dataset,aes(x=independentVar,y=dependentVar)) +
  geom_point() +
  geom_smooth(method="lm",formula=y~x)

# Use code similar to this to run a linear regression:
result = lm(dependentVar ~ independentVar,data=dataset)
# Remember, a quadratic term can be introduced in multiple ways:
result = lm(dependentVar ~ independentVar + independentVar^2,data=dataset)
result = lm(dependentVar ~ poly(independentVar,2),data=dataset)

# Adding a subject variable requires the use of 'lmer', introduced in class to do mixed effects analysis. Here is an example of how you might use lmer:
mix_result = lmer(dependentVar ~ independentVar + (1 | subjectVar),data=dataset)

# Recall also that you can do model comparisons using anova, AIC, and other measures:
AIC(result)
AIC(mix_result)
```